Microlensing as a possible probe of event-horizon structure in quasars

### 2016-07-05 paper1/ms.tex
Acknowledgement added
new figures for data-fitting section

\title{Microlensing as a possible probe of event-horizon structure in quasars}
\author[Tomozeiu et al]{Mihai Tomozeiu\thanks{mihai@physik.uzh.ch},$^{1,2}$ 
Irshad Mohammed,$^{1,2,3}$ 
Manuel Rabold,$^{1,2}$
Prasenjit Saha,$^{1,2}$  
\newauthor
and Joachim Wambsganss$^4$\\
$^1${Physik-Institut, University of Zurich, Winterthurerstrasse 190,
  8057 Zurich, Switzerland} \\
$^2${Institute for Computational Science, University of Zurich,
  Winterthurerstrasse 190, 8057 Zurich, Switzerland} \\
$^3${Theoretical Astrophysics Group, Fermi National Accelerator Laboratory, Batavia, IL 60510, USA}\\
$^4${Zentrum f\"ur Astronomie der Universit\"at Heidelberg,
  M\"onchhofstrasse 12--14, 69120 Heidelberg, Germany}
}

@@

In quasars that have been lensed by galaxies, the point-like images
sometimes show sharp brightness variations (microlensing).  These
brightness changes are associated with the innermost region of the
quasar passing behind a complicated pattern of caustics due to the
stars in the lensing galaxy.  In this paper, we study whether the
universal properties of optical caustics could enable extraction of
shape information about the central engine of quasars.  We present a
toy model with a crescent-shaped source crossing a fold caustic.  The
silhouette of a black hole over an accretion disk tends to produce
roughly crescent sources.  When a crescent source crosses a fold
caustic, the resulting light curve is noticeably different from the
case of a disc or Gaussian source.  The crescent parameters, apart
from one degeneracy, can be recovered.
\end{abstract}

@@

\begin{figure}
\centering
\includegraphics[width=0.9\hsize]{figures/IRIS567_path.eps}
\caption{\label{fig:magnification_map} Magnification map used later in
  this paper (Section~\ref{sec:numerics}). The white line marks the
  trajectory of the center of the sources for which the lightcurves
  presented in figure 12 are generated.  Note that this represents an
  atypically simple region of any realistic magnification map.}
\end{figure}

@@

\begin{figure}
\centering
        \includegraphics[width = 0.48\textwidth]{figures/M87_image.eps}
\caption{\label{fig:M87_image} The simulated image of M87's black hole
  silhouette obtained by \citep{2012MNRAS.421.1517D} from a GRMHD
  simulation presented in \citep{2009MNRAS.394L.126M}. The color bar
  is linear and scaled to reach value 1 for the maximum
  brightness.}
\end{figure}

\cite{2012MNRAS.421.1517D} have created a radiative image of M87 based on the GRMHD simulations presented in \citep{2009MNRAS.394L.126M}. The respective image reproduced in figure \ref{fig:M87_image}, has been projected to a 1D profile associated to the perpendicular direction to a fold caustic approaching the image from the right. The projection is presented in the left panel of figure 10. The amplification values of the flux of light corresponding to a microlensing event are presented in the right panel of figure 10. In general, the behaviour of the lightcurve is similar to the geometric crescent source with the caveat that the outer regions surrounding the luminous parts of the image have non-zero flux and thus are more extended than the simplified source model.

\begin{figure}
\centering
    \includegraphics[width = 0.48\textwidth]{figures/M87_shape.eps}
        \includegraphics[width = 0.48\textwidth]{figures/M87_lc.eps}
\caption{\label{fig:M87_plots} The one dimensional profile obtained
  from the numerical integration along the ordonata axis of the source
  image (figure 9) is presented in the lower panel. The corresponding
  lightcurve associated to the 1D profile is presented in the right
  panel. The points p2 and p3 mark the start and end of the equivalent
  dark disk and are associated with the moments of their overlap with
  the caustic: t2 and t3.}
\end{figure}

@@

First we generate microlensing magnification maps using the
microlensing code by \cite{1999A&A...346L...5W}.  We then generate
mock light curves for three source models --- a crescent, a uniform
disc and a Gaussian disc --- and then fit each mock light curve to all
three source models. Parameter fitting and marginalising was done by
Markov-chain Monte Carlo (MCMC).

@@

The above results suggest the following strategy for fitting a
lightcurve from an unknown source profile: first try a Gaussian-source
model; if the data reject that model, try a uniform disc; if the
uniform disc is also rejected by the data, try a crescent model.  Our
numerical experiments indicate --- assuming one of the three sources
models are correct --- that the reduced $\chi^2$ would unmask the
correct model, \textbf{and its parameters would be correctly recovered}.

@@

Furthermore, a simulated image of M87 presented in \citep{2012MNRAS.421.1517D} (figure 9) has been microlensed. On the resulting lightcurve the instances 
corresponding to the start and end of the black hole shadow and caustic overlap were distinguishable as can be seen in figure 10.

@@

\textit{Author contributions}\\
Prasenjit Saha provided the original idea and plan for the research project as well as multiple contributions to the analysis and manuscript preparation. Mihai Tomozeiu simulated and studied the ideal behaviour of the microlensing lightcurves for the different source profiles discussed and prepared the manuscript.
 Joachim Wambsganss provided the numerical code used by Manuel Rabold to create the magnification map and corresponding lightcurves 
used in the MCMC analysis performed by Irshad Mohammed in the last part of the presented work. Both Manuel Rabold and Irshad Mohammed had large contributions in writing the "Fitting Mock Data" section.

>>>>

\title{Microlensing as a possible probe of event-horizon structure in quasars}
\author[Tomozeiu et al]{Mihai Tomozeiu\thanks{mihai@physik.uzh.ch},$^{1,2}$ 
Irshad Mohammed,$^{1,2,3}$ 
Manuel Rabold,$^{1,2}$
Prasenjit Saha,$^{1,2}$  
\newauthor
and Joachim Wambsganss$^{1,4}$\\
$^1${Physik-Institut, University of Zurich, Winterthurerstrasse 190,
  8057 Zurich, Switzerland} \\
$^2${Institute for Computational Science, University of Zurich,
  Winterthurerstrasse 190, 8057 Zurich, Switzerland} \\
$^3${Theoretical Astrophysics Group, Fermi National Accelerator Laboratory, Batavia, IL 60510, USA}\\
$^4${Zentrum f\"ur Astronomie der Universit\"at Heidelberg,
  M\"onchhofstrasse 12--14, 69120 Heidelberg, Germany}
}

@@

In quasars which are lensed by galaxies, the point-like images sometimes show sharp and uncorrelated brightness 
variations (microlensing). These brightness changes are associated with the innermost region of the quasar passing 
through a complicated pattern of caustics produced by the stars in the lensing galaxy. In this paper, we study 
whether the universal properties of optical caustics could enable extraction of shape information about the central 
engine of quasars. We present a toy model with a crescent-shaped source crossing a fold caustic. The silhouette 
of a black hole over an accretion disk tends to produce roughly crescent sources. When a crescent-shaped source 
crosses a fold caustic, the resulting light curve is noticeably different from the case of a circular luminosity 
profile or Gaussian source. With good enough monitoring data, the crescent parameters, apart from one degeneracy, 
can be recovered.

\end{abstract}

@@

\begin{figure}
\centering
\includegraphics[width=0.9\hsize]{figures/IRIS567_path.eps}
\caption{\label{fig:magnification_map} Magnification map used later in
  this paper (Section~\ref{sec:numerics}). The white line marks the
  trajectory of the center of the sources for which the lightcurves
  presented in figure 11 are generated.  Note that this represents an
  atypically simple region of any realistic magnification map.}
\end{figure}

@@

\cite{2012MNRAS.421.1517D} have created a radiative image of M87 based on the GRMHD simulations presented in \citep{2009MNRAS.394L.126M}. 
The top right-most image in figure 5 of \cite{2012MNRAS.421.1517D} has been projected to a 1D profile associated to the perpendicular 
direction to a fold caustic approaching the image from the right. The projection is presented in the upper panel of figure 9. 
The amplification values of the flux of light corresponding to a microlensing event are presented in the lower panel of figure 9. 
In general, the behaviour of the lightcurve is similar to the geometric crescent source with the caveat that the outer regions 
surrounding the luminous parts of the image have non-zero flux and thus are more extended than the simplified source model.

\begin{figure}
\centering
    \includegraphics[width = 0.48\textwidth]{figures/M87_shape.eps}
        \includegraphics[width = 0.48\textwidth]{figures/M87_lc.eps}
\caption{\label{fig:M87_plots} The one dimensional profile obtained
  from the numerical integration along the ordonata axis of the source
  image is presented in the upper panel. The corresponding
  lightcurve associated to the 1D profile is presented in the lower
  panel. The points p2 and p3 mark the start and end of the equivalent
  dark disk and are associated with the moments of their overlap with
  the caustic: t2 and t3.}
\end{figure}

@@

First we generate microlensing magnification maps using the
microlensing code created by Joachim Wambsgans (\citep{1999A&A...346L...5W, 1990PhDT.......180W, 1999JCoAM.109..353W}, 
\cite{1990PhDT.......180W}, \cite{1999JCoAM.109..353W}). We then generate
mock light curves for three source models --- a crescent, a uniform
disc and a Gaussian disc --- and then fit each mock light curve to all
three source models. Parameter fitting and marginalising was done by
Markov-chain Monte Carlo (MCMC).

\begin{figure}
\centering
  \includegraphics[width=0.9\hsize]{figures/cc_forward.eps}
\caption{\label{fig:cc_forward} $\chi^2=243, P=0.32$.}
\end{figure}

\begin{figure}
\centering
  \includegraphics[width=0.9\hsize]{figures/cc_backward.eps}
\caption{\label{fig:cc_backward} $\chi^2=284, P=0.01$.}
\end{figure}

\begin{figure}
\centering
  \includegraphics[width=0.9\hsize]{figures/gc_forward.eps}
\caption{\label{fig:gc_forward} $\chi^2=344, P=3\times10^{-6}$.}
\end{figure}

@@

The above results suggest the following strategy for fitting a
lightcurve from an unknown source profile: first try a Gaussian-source
model; if the data reject that model, try a uniform disc; if the
uniform disc is also rejected by the data, try a crescent model.  Our
numerical experiments indicate --- assuming one of the three sources
models are correct --- that the reduced $\chi^2$ would unmask the
correct model, and its parameters would be correctly recovered.

@@

Furthermore, a simulated image of M87 presented in \citep{2012MNRAS.421.1517D} has been microlensed (figure 9). On the resulting lightcurve the instances 
corresponding to the start and end of the black hole shadow and caustic overlap were distinguishable.

@@

\textit{Author contributions}\\
Prasenjit Saha provided the original idea and plan for the research project as well as multiple contributions to the analysis 
and manuscript preparation. Mihai Tomozeiu simulated and studied the ideal behaviour of the microlensing lightcurves for the 
different source profiles discussed and prepared the manuscript.
Joachim Wambsganss contributed to the research planning and provided the numerical code used by Manuel Rabold to create 
the magnification map and corresponding lightcurves used in the MCMC analysis performed by Irshad Mohammed in the last 
part of the presented work. Both Manuel Rabold and Irshad Mohammed had large contributions in writing the "Fitting Mock Data" section.

\section{Acknowledgement}
IM is supported by Fermi Research Alliance, LLC under Contract No. De-AC02-07CH11359 with the United States Department of Energy.
J.W. would like to acknowledge and thank the Pauli Center for Theoretical Studies of ETH Zurich and University of Zurich for 
generous support during the Schrödinger visiting professorship in 2013.
